A Formal and Practical Metatheory of Bidirectional Type Synthesis for Simple Types

### 2023-06-12 tex/BiSig.tex
[ text ] Update the structure of paper

\title{A Formal Practical Foundation of Bidirectional Typing for Simple Types}

@@

@@

\subsubsection{Bidirectional typing}
\cite{Pierce2000}
\cite{Dunfield2021}

@@

Our assumptions and contributions:
\begin{enumerate}
  \item Syntax-directed typing rules
  \item Simple types
  \item Fully formalised in \Agda 
  \item Generic.
\end{enumerate}

\section{Using Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus as an Outline} \label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

\section{Definitions of Simple Type Theories}\label{sec:simple-type-theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$, where $\arity(\bto) = 2$ and an operation $N$ of arity $0$.
  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
\end{example}

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $\Xi$ is the number of type variables, 
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

@@

\subsection{Intrinsically Typed Terms}

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
  \label{fig:intrinsic-typing}
\end{figure}

\subsection{Raw Terms and Extrinsically Typed Terms}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{{\color{red}\rho : \Sub{\Xi}{\emptyset}} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

\section{Definitions of Bidirectional Simple Type Systems}

\subsection{Bidirectional Binding Signatures}

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi)\right)^* \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

@@

\begin{example}
  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
\end{example}

@@

\begin{theorem}[Signature annotation]

@@

\subsection{Extrinsically Typed Terms for Bidirectional Type Systems}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})^\isDir{\chk}}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\subsection{Soundness and Completeness with respect to Binding Signatures}

\subsection{Annotatability}

\begin{theorem}[Decidability for Bidirectionalisation]
  For every annotatable raw $t$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
\end{theorem}

\section{Bidirectional Type Synthesis}

\subsection{Mode Correctness}
%\subsection{Substitution Properties}
%\subsection{Generic First-Order Unification}
%\cite{McBride2003,McBride2003a}

@@

\begin{theorem}[Decidability for Bidirectional Typing Judgements]
  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode correct, the following two statements hold constructively.
  \begin{enumerate}
    \item For every context $\Gamma$ and a pre-synthesised term $\isTerm{t}$, it is decidable that there exists a type $A$ and a bidirectional typing derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
      \]
    \item For every context $\Gamma$, a pre-checked term $\isTerm{t}$, and a type $A$, it is decidable that there is a typing derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
      \]
  \end{enumerate}
\end{theorem}

\begin{theorem}[Decidability of Type Inference]
  \LT{combine the above two steps}
  Let $\Sigma$ be a type signature and $\Omega$ a binding signature with a bidirectional annotation $\Omega^{\leftrightarrows}$.
  Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
  \begin{enumerate}
    \item $t$ has typing derivation $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $\vdash_{\Sigma} A$.
    \item $t$ is not typable under $\Gamma$.
    \item there exists $t'$ such that $t$ needs more type annotations marked in $t'$ to be inferred.
  \end{enumerate}
\end{theorem}

\subsection{Decidability of Bidirectional Type Synthesis and Checking}

\begin{theorem}{Uniqueness of Synthesised Types}\label{thm:unique-syn}

>>>>

\title{A Formal and Practical Metatheory of Bidirectional Type Synthesis for Simple Types}

@@

\begin{enumerate}
  \item Type checking as a transition from (parsed) abstract syntax trees to well-typed terms.
  \item Bidirectional typing is the only solution that works that scales well.
  \item Not yet a general theory but by now a design principle.
    No `type checker generator' as each type checker is still designed and implemented on a case-by-case basis, but can be derived from a syntax-directed type theory `easily'.
  \item The existing work of bidirectional typing does not take account of missing annotations carefully enough. 
  \item A meta-problem: existing work is based on (or at least left unspecified) classical logic.
    One has to state soundness of completeness of their type checking algorithm in a less straightforward way.
  \item A review of what has been done towards this line of research.
  \item Finally, our responses (contributions of this paper).
\end{enumerate}

@@

\subsubsection{Bidirectional typing}

\cite{Pierce2000}
\cite{Dunfield2021}

@@

Our assumptions and contributions:
\begin{enumerate}
  \item Assumption: Syntax-directed typing rules
  \item Practical: bidirectionalisation over annotatability, decidability over soundness and completeness.
  \item Generic: Type checking once and for all bidirectional type systems for simple types.
  \item Correct and Constructive: Fully formalised in Martin-Löf type theory (with Axiom K).
    LEM is not assumed, so a decidable statement has both computational and non-trivial logical readings.
\end{enumerate}

Plan of this paper:
\begin{enumerate}
  \item Key ideas (\Cref{sec:key-ideas})
  \item Definitions for simple type theories and for bidirectional type systems (signature erasure and mode annotation) (\Cref{sec:defs})
  \item Soundness, completeness, bidirectionalisation, and annotatability (\Cref{sec:annotatability})
    \Josh{Erasure from bidirectional typing derivation to raw terms with a mode}
  \item Bidirectional type inference (\Cref{sec:type-synthesis})
  \item Formalisation with further examples (\Cref{sec:formalisation})
  \item Future work (\Cref{sec:future})
\end{enumerate}
  \LT{better title?}

\section{Core Ideas with Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus} \label{sec:key-ideas}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}\;(\text{app})
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}\;(\text{abs})
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

\section{Definitions for Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}

\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$ and a base type (nullary operation) $N$.
  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
\end{example}

\subsection{Binding Signatures} \label{subsec:binding-sig}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $\Xi$ is the number of type variables, 
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

@@

%\subsection{Intrinsically Typed Terms}
%
%Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\Gamma \vdash_{\Sigma, \Omega} A$ constructed with only the rule $(\var)$ and the rule scheme $(\tmOp)$ displayed in \Cref{fig:intrinsic-typing}.
%
%The substitution $\rho\colon \Sub{\Xi}{\emptyset}$ is used to instantiate variables in the local context $\Xi$ with concrete types.
%Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\tmOp$ to ensure all types are well-formed without any use of type variables.
%
%\begin{figure}
%  \centering
%  \small
%  \begin{mathpar}
%    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
%    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
%    \and
%    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
%    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
%    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
%  \end{mathpar}
%  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
%  \label{fig:intrinsic-typing}
%\end{figure}

\subsection{Simple Type Theories}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{\rho : \Sub{\Xi}{\emptyset} \\ \Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems}

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times \isDir{\Mode}.
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, A_1, d_1\right), \ldots, \left(\Delta_{n}, A_{n}, d_n\right) \right], A, d\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

@@

\begin{example}
  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{subsec:binding-sig} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
\end{example}

@@

\LT{Signature Erasure}
\begin{theorem}[Signature annotation]

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}^\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}^\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}^\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})^\isDir{\chk}}\;(\text{sub})
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}^\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}^\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}^\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms with a mode $d$ for a bidirectional type system $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}^\isDir{d}$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\section{Soundness, Completeness, and Bidirectionalisation} \label{sec:annotatability}

\subsection{Soundness and Completeness}

\subsection{Bidirectionalisation}
\begin{theorem}[Decidability for Bidirectionalisation] \label{thm:bidirectionalisation}
  For every raw term $t$ and a mode $d$, it is decidable that $t$ has a mode $d$, i.e.\ $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t}\,\isDir{d}$ is derivable.
\end{theorem}

\subsection{Completeness versus Annotatability}

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}

\subsection{Mode Correctness}

\begin{theorem}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
  Let $(\Sigma, \Omega)$ be a bidirectional type system where $\Omega$ is mode-correct.
  Then, for every raw $t$ with a synthesis mode $\Rightarrow$, and any derivations of 
  \[
    \Gamma \vdash t \Rightarrow A
    \quad\text{and}\quad
    \Gamma \vdash t \Rightarrow B
  \]
  for some types $A$ and $B$, the synthesised types must be equal, i.e.\ $A = B$.
\end{theorem}

@@

\begin{theorem}[Decidability for Bidirectional Typing Synthesis] \label{thm:bidirectional-type-synthesis}
  For every type signature $\Sigma$ and bidirectional binding signature $\Omega$ for $\Sigma$ that is mode-correct, the following two statements hold.

  \begin{enumerate}
    \item For every context $\Gamma$ and a raw term $\isTerm{t}$ with a synthesis mode, it is decidable that there exists a type $A$ and a derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A.
      \]
    \item For every context $\Gamma$, a raw term $\isTerm{t}$ with a checking mode, and a type $A$, it is decidable that there is a derivation of
      \[
        \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A.
      \]
  \end{enumerate}
\end{theorem}

\begin{corollary}[`Decidability' of Type Synthesis]
  \LT{Still call it `decidability'?}
  Let $\Sigma$ be a type signature and $\Omega$ a bidirectional binding signature.
  Then, for every context $\Gamma$ and raw term $t$, one and only one of the following statements holds:
  \begin{enumerate}
    \item there is a typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for some type $A$.
    \item there is no typing derivation of $\Gamma \vdash_{\Sigma, \Omega} t : A$ for any type $A$.
    \item $t$ cannot have a synthesis mode.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{thm:bidirectionalisation} and \Cref{thm:bidirectional-type-synthesis}.

\end{proof}

\subsection{Proof for Decidability of Bidirectional Type Synthesis}
